A general Nullstellensatz for generalized spaces

### 2019-03-18 paper-qcoh.tex
Started section on applications

\section{Introduction}

@@

In time, further nongeometric sequents holding in the big Zariski topos of an
arbitrary base scheme have been found~\cite[Section~18.4]{blechschmidt:phd}. These include:
\begin{itemize}
\item $\affl$ is \emph{anonymously algebraically closed} in the sense that
any monic polynomial~$p \? \affl[T]$ of degree at least one does \notnot have a
zero. \smallskip
\item The Nullstellensatz holds:
Let $f_1,\ldots,f_m \in \affl[X_1,\ldots,X_n]$ be polynomials without a common
zero in~$(\affl)^n$. Then there are polynomials~$g_1,\ldots,g_m \in
\affl[X_1,\ldots,X_n]$ such that~$\sum_i g_i f_i = 1$. \smallskip
\item Any function~$\affl \to \affl$ is given by a unique polynomial. \smallskip
\item $\affl$ is microaffine: Let~$\Delta = \{ \varepsilon \?
\affl \,|\, \varepsilon^2 = 0 \}$. Let~$f : \Delta \to \affl$ be an arbitrary
function. Then there are unique elements~$a, b \? \affl$ such
that~$f(\varepsilon) = a + b \varepsilon$ for all~$\varepsilon \? \Delta$. \smallskip
\item $\affl$ is \emph{synthetically quasicoherent}: For any
finitely presentable~$\affl$-algebra~$A$, the canonical homomorphism~$A \to
(\affl)^{\Spec(A)}$, where~$\Spec(A)$ is defined as the set
of~$\affl$-algebra homomorphisms~$A \to \affl$, is bijective.
\end{itemize}
All of these nongeometric sequents are useful for the purposes of synthetic
algebraic geometry, the desire to carry out algebraic geometry in a language
close to the simple language on the 19th and the beginning of the 20th century
while still being fully rigorous and fully general, working over arbitrary base
schemes instead of restricting to the field of complex numbers.
\medskip

@@

@@

\begin{thm}The generic~$\TT$-model is \emph{quasicoherent} in the following sense:
From the point of view of~$\Set[\TT]$, for any finitely
presentable~$U_\TT$-algebra~$A$ (finitely presented object in the category
of~$U_\TT$-algebras), the canonical~$U_\TT$-algebra homomorphism
\[ A \longrightarrow (U_\TT)^{\Spec(A)} \]
is an isomorphism.
\end{thm}

@@

@@

>>>>

\section{Introduction}
\label{sect:introduction}

@@

In time, further nongeometric sequents holding in the big Zariski topos of an
arbitrary base scheme have been found~\cite[Section~18.4]{blechschmidt:phd}. These include:
\begin{itemize}
\item $\affl$ is \emph{anonymously algebraically closed} in the sense that
any monic polynomial~$p \? \affl[T]$ of degree at least one does \notnot have a
zero. \smallskip
\item The Nullstellensatz holds:
Let $f_1,\ldots,f_m \in \affl[X_1,\ldots,X_n]$ be polynomials without a common
zero in~$(\affl)^n$. Then there are polynomials~$g_1,\ldots,g_m \in
\affl[X_1,\ldots,X_n]$ such that~$\sum_i g_i f_i = 1$. \smallskip
\item Any function~$\affl \to \affl$ is given by a unique polynomial. \smallskip
\item $\affl$ is microaffine: Let~$\Delta = \{ \varepsilon \?
\affl \,|\, \varepsilon^2 = 0 \}$. Let~$f : \Delta \to \affl$ be an arbitrary
function. Then there are unique elements~$a, b \? \affl$ such
that~$f(\varepsilon) = a + b \varepsilon$ for all~$\varepsilon \? \Delta$. \smallskip
\item $\affl$ is \emph{synthetically quasicoherent}: For any
finitely presentable~$\affl$-algebra~$A$, the canonical homomorphism~$A \to
(\affl)^{\Spec(A)}$, where~$\Spec(A)$ is defined as the set
of~$\affl$-algebra homomorphisms~$A \to \affl$, is bijective.
\end{itemize}
% Take care: In sect:applications, we state that each of these nongeometric
% sequents also holds in the ring classifier.
All of these nongeometric sequents are useful for the purposes of synthetic
algebraic geometry, the desire to carry out algebraic geometry in a language
close to the simple language of the 19th and the beginning of the 20th century
while still being fully rigorous and fully general, working over arbitrary base
schemes instead of restricting to the field of complex numbers.
\medskip

@@

The purpose of this section is to redo the development in the special case of
Horn theories. XXX

@@

\begin{thm}\label{thm:quasicoherence}
The generic~$\TT$-model is \emph{quasicoherent} in the following sense:
From the point of view of~$\Set[\TT]$, for any finitely
presentable~$U_\TT$-algebra~$A$ (finitely presented object in the category
of~$U_\TT$-algebras), the canonical~$U_\TT$-algebra homomorphism
\[ A \longrightarrow (U_\TT)^{\Spec(A)} \]
is an isomorphism.
\end{thm}

% XXX justification name quasicoherent

@@

% XXX subtopos case

@@

\section{Nongeometric sequents in the ring classifier}

Each of the nongeometric sequents mentioned in Section~\ref{sect:introduction}
can be deduced from the first-order or higher-order Nullstellensatz. Here we
briefly indicate how this works using two examples.

Let~$\TT$ be the theory of rings. Then, internally to~$\Set[\TT]$, we have for
any element~$x \? U_\TT$:
\begin{align*}
  (x = 0 \Rightarrow 1 = 0) &\Longrightarrow \speak{$x$ is invertible} \quad\text{and} \\
  (\speak{$x$ is invertible} \Rightarrow 1 = 0) &\Longrightarrow \speak{$x$ is nilpotent}.
\end{align*}
To verify the first claim, assume~$(x = 0 \Rightarrow 1 = 0)$. By the
Nullstellensatz of Theorem~\ref{thm:nullstellensatz}, the
theory~$\ul{\TT}/U_\TT$ proves this fact. Hence in particular it holds in
the~$U_\TT$-algebra~$U_\TT/(x)$. Because~$x = 0$ in~$U_\TT/(x)$, we have~$1 =
0$ in~$U_\TT/(x)$, that is~$1 \in (x)$, hence~$x$ is invertible.

To verify the second claim, assume~$(\speak{$x$ is invertible} \Rightarrow 1 =
0)$. By the Nullstellensatz, the theory~$\ul{\TT}/U_\TT$ proves this fact,
hence in particular it holds in~$U_\TT[x^{-1}]$. Because~$x$ is invertible in
this~$U_\TT$-algebra, we have~$1 = 0$ in~$U_\TT[x^{-1}]$. Hence~$x$ is
nilpotent.

Thanks to the version of the Nullstellensatz pertaining Horn consequents given
in Theorem~\ref{thm:nullstellensatz-horn}, these two arguments can also be
carried out in the classifying topos of local rings. It is instructive to
consider how we would have to argue if we wanted to use only the unmodified
Nullstellensatz:

Assume~$(x = 0 \Rightarrow 1 = 0)$. By the Nullstellensatz, the theory of local
rings proves this fact. From this we may not conclude that it holds
for~$U_\TT/(x)$, as this~$U_\TT$-algebra might not be local. But we may
conclude that it holds for~$\O_{\Spec(U_\TT/(x))}$, the structure sheaf of the affine scheme
given by~$U_\TT/(x)$. This sheaf is not a model of~$\ul{\TT}/U_\TT$, but it is
so from the point of view of the topos of sheaves over~$\Spec(U_\TT/(x))$. From
the point of view of that topos,~$x = 0$ in~$\O_{\Spec(U_\TT/(x))}$, hence~$1 =
0$ in~$\O_{\Spec(U_\TT/(x))}$. This amounts to~$U_\TT/(x)$ being the zero ring,
hence~$x$ is invertible in~$U_\TT$.

\section{On the Kock--Lawvere axiom in synthetic differential geometry}

The generic local ring~$\affl$ validates the following strong form of the Kock--Lawvere
axiom (Theorem~\ref{thm:quasicoherence}): \emph{For any finitely
presented~$\affl$-algebra~$A$, the canonical map~$A \to (\affl)^{\Spec(A)}$ is
an isomorphism of~$\affl$-algebras.} The special case~$A \defeq \affl[X]$
implies that any map~$\affl \to \affl$ is given by a polynomial,
since~$\Spec(\affl[X]) = \Hom_\affl(\affl[X],\affl) \cong \affl$.

In contrast, in most models of synthetic differential geometry, we only have
the following weaker statement: \emph{For any Weil algebra~$A$, the canonical
map~$A \to \RR^{\Spec(A)}$ is an isomorphism of~$\RR$-algebras.} Here~$\RR$
refers to the canonical line object (the image of the manifold~$\RR^1$ in the
model) and~$\Spec(A)$ is the set of~$\RR$-algebra homomorphisms~$A \to \RR$.
Any Weil algebra is finitely presented, but the converse fails; in particular,
the polynomial algebra~$\RR[X]$ is not a Weil algebra, hence there is no reason
to believe that any map~$\RR \to \RR$ is given by a polynomial.

The quasicoherence statement of
Theorem~\ref{thm:quasicoherence} can shed light on this discrepancy. % XXX
